---
title: "Titanic"
author: "Adrien Giraud"
date: "22 November 2015"
output: html_document
---
# 1. Working environment preparation and database loading

## 1.1. Libraries and working directory setting
```{r warning=FALSE, message =FALSE}
# Load libraries :
library(data.table)
library(ggplot2)
library(stats)
library(randomForest)
library(Hmisc)
library(Amelia)
library(party)
library(methods)
library(rpart)
library(gbm)
my_path <- "/Users/adrien/Documents/axa/titanic"
setwd(my_path)
```
## 1.2. Database loading
```{r  warning=FALSE }
missing.types <- c("NA", "")
train.column.types <- c("integer",   # PassengerId
                        "factor",    # Survived 
                        'factor',    # Pclass
                        'character', # Name
                        'factor',    # Sex
                        'numeric',   # Age
                        'integer',   # SibSp
                        'integer',   # Parch
                        'character', # Ticket
                        'numeric',   # Fare
                        'character', # Cabin
                        'factor'     # Embarked
)

test.column.types <- c('integer',   # PassengerId
                        'factor',    # Pclass
                        'character', # Name
                        'factor',    # Sex
                        'numeric',   # Age
                        'integer',   # SibSp
                        'integer',   # Parch
                        'character', # Ticket
                        'numeric',   # Fare
                        'character', # Cabin
                        'factor'     # Embarked
)
train<-fread("train.csv",sep=",",colClasses=train.column.types,na.strings=c("NA", ""),header=T)
test<-fread("test.csv",sep=",",colClasses=test.column.types,na.strings=c("NA", ""),header=T)
```

# 2. Descriptive statistics

## 2.1. Missing data checking 
```{r  warning=FALSE}
missmap(train, col=c("yellow", "blue"), legend = FALSE, main = "Training set")
missmap(test, col=c("yellow", "blue"), legend = FALSE, main = "Test set")
check.missing<-function(x) return(paste0(round(sum(is.na(x))/length(x),4)*100,'%'))
data.frame(sapply(train,check.missing))
data.frame(sapply(test,check.missing))
```
Observations :

* roughly 20 percent of the Age data is missing
* well above 70 percent of the passengers cannot be linked to a specific cabin number. 

Conclusions :

* the proportion of Age "missings" is likely small enough for reasonable replacement with some form of imputation
* the cabin missings seem too extensive to make reliable imputation possible. 

## 2.2. General data description
```{r}
barplot(table(train$Survived),
        names.arg = c("Perished", "Survived"),
        main="Survived (passenger fate)",ylab="nb of passengers",col="grey")

barplot(table(train$Pclass), 
        names.arg = c("first", "second", "third"),
        main="Passenger traveling class distribution",ylab="nb of passengers",col="grey")

barplot(table(train$Sex), main="Sex (gender)",ylab="nb of passengers",col="grey")

hist(train$Age, main="Age distribution", xlab = "year",col="grey")

barplot(table(train$Embarked), names.arg = c("Cherbourg", "Queenstown", "Southampton"),main="Embarked (port of embarkation)", col="grey")
```

Observations :

* more passengers perished than survived
* about twice as many passengers in 3rd class than in either 1st or 2nd
* male passengers far outnumbered females
* most passangers joined the vessel at Southampton



## 2.3. Survival rate
```{r}
mosaicplot(train$Sex ~ train$Survived, main="Passenger Survival by Gender",
           color=c("#8dd3c7", "#fb8072"), shade=FALSE,  xlab="", ylab="",
           off=c(0), cex.axis=1.4)

mosaicplot(train$Pclass ~ train$Survived, main="Passenger Survival by Class",
           color=c("#8dd3c7", "#fb8072"), shade=FALSE,  xlab="", ylab="",
           off=c(0), cex.axis=1.4)

mosaicplot(train$Embarked ~ train$Survived, main="Passenger Survival by port of embarkation",
           color=c("#8dd3c7", "#fb8072"), shade=FALSE,  xlab="", ylab="",
           off=c(0), cex.axis=1.4)

```

The following mosaic suggests that :

* gender should certainly prove to be a prominent feature in the final model
* traveling class did influence the odds of a passenger's survival
* passenger survival didn't vary as much across the three ports of embarkation as it did between genders and traveling classes

# 3. Data cleaning
## 3.1. Combine train/test data for pre-processing
```{r}
train$Cat<-'train'
test$Cat<-'test'
test$Survived<-NA
full<-rbind(train,test)
```
##3.2. Feature engineering
### 3.2.1. Embarked feature
```{r} 
sum(is.na(full$Embarked))
full$Embarked[is.na(full$Embarked)]<-'S'
```
It reveals just two missings.It should be fine to replace those missings with "S", the most common value.

### 3.2.2. Title feature

```{r} 

full$Title <- sapply(full$Name, FUN=function(x) {strsplit(x, split='[,.]')[[1]][2]})
full$Title <- sub(' ', '', full$Title)
full$Title[full$Title %in% c('Mme', 'Mlle')] <- 'Mlle'
full$Title[full$Title %in% c('Capt', 'Don', 'Major', 'Sir')] <- 'Sir'
full$Title[full$Title %in% c('Dona', 'Lady', 'the Countess', 'Jonkheer')] <- 'Lady'
full$Title <- factor(full$Title)

```

### Fare Feature 
```{r} 
fit.Fare<-rpart(Fare[!is.na(Fare)]~Pclass+Title+Sex+SibSp+Parch,data=full[!is.na(full$Fare),],method='anova')
```
### 3.2.3. Age feature

#### First approach

A common approach to this type of situation is to replacing the missings with the average (or the median) of the available values.
```{r} 
mean(full$Age,na.rm=T)
median(full$Age,na.rm=T)
```
In this case, that would mean replacing 177 missing Age values with 29.9 (or 28). Taking that approach would be fine if only a small fraction of the ages were missing. However, with missings accounting for 20 percent of all Age data in a relatively small data set (<900 records), one could justify a search for a more refined method of imputation. 

#### Second approach
We make a prediction of a passengers Age using the other variables and a decision tree model

```{r}
#feature creation 
full$FamilySize <- full$SibSp + full$Parch + 1

predicted_age <- rpart(Age ~ Pclass + Sex + SibSp + Parch + Fare + Embarked + Title + FamilySize,
                       data=full[!is.na(full$Age),], method="anova")
full$Age[is.na(full$Age)] <- predict(predicted_age, full[is.na(full$Age),])
```

### 3.2.4. Miscellaneous  feature engineering

```{r}

#Adding Mother
full$Mother<-0
full$Mother[full$Sex=='female' & full$Parch>0 & full$Age>18 & full$Title!='Mlle']<-1

#Adding Child
full$Child<-0
full$Child[full$Parch>0 & full$Age<=18]<- 1

#Exact Deck from Cabin number
full$Deck<-sapply(full$Cabin, function(x) strsplit(x,NULL)[[1]][1])

#Excat Position from Cabin number
full$CabinNum<-sapply(full$Cabin,function(x) strsplit(x,'[A-Z]')[[1]][2])
full$num<-as.numeric(full$CabinNum)
num<-full$num[!is.na(full$num)]
Pos<-kmeans(num,3)
full$CabinPos[!is.na(full$num)]<-Pos$cluster
full$CabinPos<-factor(full$CabinPos)
levels(full$CabinPos)<-c('Front','End','Middle')
full$num<-NULL

Surname<-sapply(full$Name,function(x) strsplit(x,'[.,]')[[1]][1])
FamilyId<-paste0(full$FamilySize,Surname)
full$FamilyId<-factor(FamilyId)
Family<-data.frame(table(FamilyId))
SmallFamily<-Family$FamilyId[Family$Freq<=2]
FamilyId[FamilyId %in% SmallFamily]<-'Small'
full$FamilyId2<-factor(FamilyId)
```

###split train/test data
```{r}
full<-transform(full,
                Pclass=factor(Pclass),
                Sex=factor(Sex),
                Embarked=factor(Embarked),
                Title=factor(Title),
                Mother=factor(Mother),
                Child=factor(Child),
                Deck=factor(Deck),
                FamilyId=factor(FamilyId),
                FamilyId2=factor(FamilyId2),
                Survived=factor(Survived)
                )


train<-full[full$Cat=='train',]
test<-full[full$Cat=='test',]

#randomForest method, (not support variables with too many levels, e.g. FamilyId here)
#fit.rf<-randomForest(Survived~Pclass+Sex+Age+SibSp+Parch+Fare+Embarked+Title+Mother+Child+Deck,data=train,ntree=1250,importance=T)
#plot(fit.rf,main='randomForest error rate')
#imp<-importance(fit.rf,type='1')
#imp<-imp[order(imp),]
#(imp)
#varImpPlot(fit.rf, main="randomForest - variable Importance")
#
#write submission
#test$Survived<-predict(fit.rf,test,OOB=TRUE,type='response')
#submission<-test[,1:2]
#write.csv(submission,'submission_randomForest.csv',row.names=F)


#cforest (conditional inference tree) method, (support variables with more levels and missing values, with unbiased prediction)
#fit.cf<-cforest(Survived~FamilyId2+CabinPos+Deck+Pclass+Sex+Age+SibSp+Parch+Fare+Embarked+Title+Mother+Child+Deck,data=train,controls=cforest_unbiased(ntree=500, mtry=3))
#test$Survived<-predict(fit.cf,test,OOB=TRUE,type='response')
#submission<-test[,1:2]
#write.csv(submission,'submission_cforest.csv',row.names=F)


fit.gbm<-gbm(Survived~FamilyId2+CabinPos+Deck+Pclass+Sex+Age+SibSp+Parch+Fare+Embarked+Title+Mother+Child+Deck,
             data=train,
             distribution="gaussian",
             n.trees=1000,
             shrinkage=0.005,
             bag.fraction=0.7,
             interaction.depth=3)

preds<-predict(fit.gbm,test,OOB=TRUE,type='response',n.trees=1000)
normPreds<-(preds-min(preds))/(max(preds)-min(preds))
test$Survived<-as.numeric(normPreds > 0.5)
submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
write.csv(submit,'submission_gbm.csv',row.names=F)


```


